\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-11-09 doc/publications/plos16/plos16.tex
added conclusion

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R terms from the {\sc randomForest} package
\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate the
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
\cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a random forest model to
  have a very poor
\marginpar{is this true. If unsupported, remove it. Rob}
  fit, and be useless for prediction, and still have a useful ranking of variable importance;
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth}
    to 4 in their classification example. This is wrong. The
    individual trees should be grown to a large size to ensure that
    they give an unbiased estimate.}.
  In the TCGA example the RandomForests code produced trees with depths of 30+. 
\end{itemize}

@@

\marginpar{figure needs bigger text}
% \begin{figure}[tbhp] 
%  \begin{adjustwidth}{-1.00in}{0in}
%   \begin{subfigure}[b]{0.5\linewidth}
%     \centering
%     \includegraphics[totalheight=6cm]{./figs/var_oder.png} 
%     \caption{Variable importance of top 20 variables.} 
%     \label{fig7:a} 
%     \vspace{4ex}
%   \end{subfigure}%% 
%   \begin{subfigure}[b]{0.5\linewidth}
%     \centering
%     \includegraphics[totalheight=6cm]{./figs/var_oob.png}
%     \caption{OOB classification error} 
%     \label{fig7:b} 
%     \vspace{4ex}
%   \end{subfigure} 
% \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
%     forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
%   importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
%   aggregated importance of variables not present in the linear model. The dotted line represents the original linear
%   model coefficients.
%  {\bf b)} OOB (out of bag) classification error
%   estimate as function of \mtry\ fraction.}
%   \label{figure:synth}
% \end{adjustwidth}
% \end{figure}

@@

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l|l|l|l|l|l|p{1cm}|}
\hline
\bf{Genome \%}                      & \bf{Method} & \bf{Error}  & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077  &                  &                                                                   \\
                                                & Spark ML  &            &                  &                                                                   \\
                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger - R      & 0.06       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger - c++     & 0.06       & 7 mins           & --nthreads 20                                                     \\

 \hline
\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  &                                                                \\
                                                    & RF - R        & NA         & NA               & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface} \\
                                                    & H2O           & 0.05       & 16 hours         & 600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
                                                    & ranger - R    & 0.06       & 4 1/2 mins       & 21GB (one processor) \\
\hline
\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min        & 4GB \\
                                                    & Spark ML  & 0.94       & 8hr 29min        & 8GB \\
                                                    & ranger - R      &  0.05       & 28 mins          & num.threads=10\footnote{don't
                                                                                                      know how to get the memory
                                                                                                      used,   This was done on dpas-03} \\
\hline
\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min                                             & 4GB \\
                                                    & RF - Spark ML  &            &                                                      & \\
                                                    & ranger - R      & 0.03      & 58 mins\footnote{ but took 6 hours to load the
                                                                                   data}
                                                                                                                                         &
                                                                                                                                           909GB\footnote{I
                                                                                                                                           think
                                                                                                                                           this
                                                                                                                                           is
                                                                                                                                           total
                                                                                                                                           memory. The
                                                                                                                                           job
                                                                                                                                           was
                                                                                                                                           run
                                                                                                                                           with
                                                                                                                                                   ntasks-per-node=160
                                                                                                                                           cores-per-socket=10
                                                                                                                                           on
                                                                                                                                           ruby
                                                                                                                                           }  \\
& ranger - R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\
\hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX (0.96) & 7hr 14min & 8GB \\
& Spark ML & - & - & - \\
& ranger (C++)       &        NA     &        NA     &            80 CPU -- runs out of memory (12.8GB per CPU) \\
& ranger (C++)       &        NA     &        NA     &            160
                                                        CPU -- runs
                                                        out of time at
                                                        20 hours \\
\hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{minipage}
\end{table}

@@

@@

%  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
%    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
%  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
%    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691

%http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
 %http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
%overcomes this issue and is hence ideally suited to this application.  Furthermore, random forests are able to capture interactions
%between features, which is of importance for modelling complex polygenic diseases.

% We also investigated classifying the samples using random forests.  We chose random forests due to its ability to
% capture complex interactions between features. Also, due to it being an ensemble method where trees can be formed
% independently, which, is ideal for parallelization. Furthermore, the ensemble of weak learners can help to minimise
% overfitting and variance.

% The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a
% huge number of samples with a modest dimensionality, the 1000 Genomes data instead contains a huge number of
% dimensions (features). As a result, the default random forests implementation is unable to partition the data into
% small-enough tasks for nodes on the computer cluster to handle.

% For figure citations, please use "Fig." instead of "Figure".
%Nulla mi mi, Fig.~\ref{fig1}  Fusce fringilla erat porttitor lectus cursus, \nameref{S1_Video} vel sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.
%\begin{figure}[h]
%\caption{{\bf Figure Title first bold sentence Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}
%Figure Caption Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
%\label{fig1}
%\end{figure}
%\begin{enumerate}
%\item{react}
%\item{diffuse free particles}
%\item{increment time by dt and go to 1}
%\end{enumerate}

% Results and Discussion can be combined.

%\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
%\caption{
%{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
%\begin{tabular}{|l|l|l|l|l|l|l|l|}
%\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
%$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
%$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
%$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
%\end{tabular}
%\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
%\end{flushleft}
%\label{table1}
%\end{adjustwidth}
%\end{table}

% \begin{figure}[tbph]
% \begin{adjustwidth}{-1.00in}{0in}
% \begin{tabular}{ll}
% a)& b)\\
% \includegraphics[totalheight=6cm]{./figs/var_oder.png} & 
% \includegraphics[totalheight=6cm]{./figs/var_oob.png} \\
% \end{tabular}
% \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
%     forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
%   importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
%   aggregated importance of variables not present in the linear model. The dotted line represents the original linear
%   model coefficients.
%  {\bf b)} OOB classification error
%   estimate as function of \mtry\ fraction.}
% \label{figure:synth}
% \end{adjustwidth}
% \end{figure}

% \marginpar{can we remove this table and just rely on the figure?}
% \begin{table}[!ht]
% \centering
% \caption{
% {\bf Time in seconds to build a random forest of 100
%   trees with specified \mtry\ fraction using given number of CPU cores.}}
% \begin{tabular}{|l|l|l|l|l|l|}
% \hline
% \multicolumn{2}{|l|}{\multirow{2}{*}{}}             & \multicolumn{4}{c|}{Number of CPU cores} \\
% \cline{3-6}
% \multicolumn{2}{|l|}{}                              & \bf{16} & \bf{32} & \bf{64} & \bf{128} \\
% \hline
% \multirow{4}{*}{\mtry\ (fraction)}        & \bf{0.10} & 708.3   & 340.0   & 186.0   & 134.5 \\
%                                         & \bf{0.25} & 1134.9  & 535.0   & 270.8   & 179.6 \\
%                                         & \bf{0.50} & 1724.1  & 856.6   & 428.7   & 243.8 \\
%                                         & \bf{0.80} & 2313.5  & 1125.7  & 569.1   & 364.6 \\
% \hline
% \end{tabular}
% \begin{flushleft} 
% \end{flushleft}
% \label{table10}
% \end{table}

>>>>

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R terms from the {\sc randomForest} package
\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate the
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a very poor fit and still have a useful ranking of variable
  importance;
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth}
    to 4 in their classification example. This is wrong. The
    individual trees should be grown to a large size to ensure that
    they give an unbiased estimate.}.
  In the TCGA example the RandomForests code produced trees with depths of 30+. 
\end{itemize}

@@

\marginpar{figure needs bigger text}
% \begin{figure}[tbhp] 
%  \begin{adjustwidth}{-1.00in}{0in}
%   \begin{subfigure}[b]{0.5\linewidth}
%     \centering
%     \includegraphics[totalheight=6cm]{./figs/var_oder.png} 
%     \caption{Variable importance of top 20 variables.} 
%     \label{fig7:a} 
%     \vspace{4ex}
%   \end{subfigure}%% 
%   \begin{subfigure}[b]{0.5\linewidth}
%     \centering
%     \includegraphics[totalheight=6cm]{./figs/var_oob.png}
%     \caption{OOB classification error} 
%     \label{fig7:b} 
%     \vspace{4ex}
%   \end{subfigure} 
% \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
%     forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
%   importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
%   aggregated importance of variables not present in the linear model. The dotted line represents the original linear
%   model coefficients.
%  {\bf b)} OOB (out of bag) classification error
%   estimate as function of \mtry\ fraction.}
%   \label{figure:synth}
% \end{adjustwidth}
% \end{figure}
\marginpar{this was done with n=100000. Needs to redone with n=2.5M}

@@

\begin{table}[!ht]
\begin{minipage}{\textwidth}
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l|l|l|l|l|l||}
\hline
\bf{Genome \%}                      & \bf{Method} & \bf{Error}  & \bf{Runtime}  \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077  &                         \\
                                                & Spark ML  &  Aidan           &           Aidan         \\
                                                & RF - R         & 0.05       & 24 min           \\
                                                & ranger - R      & 0.06       & 1 min 10 seconds       \\
                                                & ranger - c++     & 0.06       & 7 mins            \\

 \hline
\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  \\
                                                    & RF - R        & NA         & NA                \\
                                                    & H2O           & 0.05       & 16 hours         \\
                                                    & ranger - R    & 0.06       & 4 1/2 mins     \\
\hline
\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min         \\
                                                    & Spark ML  & 0.94       & 8hr 29min        \\
                                                    & ranger - R      &  0.05       & 28 mins         \\
\hline
\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min               \\
                                                    & RF - Spark ML  &    Aidan         &   Aidan               \\
                                                    & ranger - R      & 0.03      &         \\
& ranger - R        &             0.973     &      03:58           \\
\hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX  & 7hr 14min \\
& Spark ML & Aidan & Aidan  \\
& ranger (C++)       &        NA     &        NA      \\
& ranger (C++)       &        NA     &        NA    \\
\hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{table1}
%\end{adjustwidth}
\end{minipage}
\end{table}

@@

We have demonstrated that using a different model of paralleization can extend the random forest to delaing with the
case of an extremely large number of variables. By comparing this implementation to other implentations (including those
optimized for large data sets) we have demonstrated the utility of this implentation.

We have treated the case of variable selection in a $p >> n$ model where most of the variables are uninformative. We
have also demostrated the utility of the model for large GWAs data sets.

\marginpar{signal rather than ``significant variables''}

Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result based on arguments from
combinatorial geometry. This has applications in a large number of areas including variable selection in high
diensions. Consider figure \ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover
the important variables, plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant
variables). There is a distinct doundary, shown empirically and also based on arguments from combinatorial geometry, to the
region where we can reliably recover significant variables.

We note that the examples we are considering lie on the extreme left of the plot, close to the origin. So there are
limits to what can be recovered in the presence of noise. However in the case of data that is both big and wide, \cursedforest and
other VariantSpark methods may provide a useful tool.

\marginpar{redo plot}

\begin{figure}[tbhp]
\includegraphics[totalheight=6cm]{./figs/phase-diagram-equivalence.png} 
\caption{{\bf The phase chage diagram.}}
\label{figure:phase-diagram-equivalence.png} 
\end{figure}

@@

### 2016-11-09 doc/publications/plos16/supplementary_information.tex
added conclusion

>>>>

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l|l|l|l|l|l|p{1cm}|}
\hline
\bf{Genome \%}                      & \bf{Method} & \bf{Error}  & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077  &                  &                                                                   \\
                                                & Spark ML  &            &                  &                                                                   \\
                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger - R      & 0.06       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger - c++     & 0.06       & 7 mins           & --nthreads 20                                                     \\
                                                    & ranger - R    & 0.06       & 4 1/2 mins       & 21GB (one processor) \\
\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min        & 4GB \\
                                                    & Spark ML  & 0.94       & 8hr 29min        & 8GB \\
                                                    & ranger - R      &  0.05       & 28 mins          & num.threads=10\footnote{don't
\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min                                             & 4GB \\
                                                    & ranger - R      & 0.03      & 58 mins\footnote{ but took 6 hours to load the

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX (0.96) & 7hr 14min & 8GB \\
& Spark ML & - & - & - \\
& ranger (C++)       &        NA     &        NA     &            80 CPU -- runs out of memory (12.8GB per CPU) \\
& ranger (C++)       &        NA     &        NA     &            160
                                                        CPU -- runs
                                                        out of time at
                                                        20 hours \\
\hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{minipage}
\end{table}
